---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# ssdinversionrr

<!-- badges: start -->
<!-- badges: end -->

This is a research compendium which is an R package for reproducible research that includes package data, package functions, package scripts, as well as complete analysis scripts and a markdown file for reproducing the research in its entirety. 

The analyses examines the issue of the inversion principle in model averaging of statistical distributions focusing on two differing case studies, the species sensitivity distribution and estimation of community protection values, as well as analysis of extreme river discharge events.

```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub.
